\section{Helper functions}
Helper functions are provided in {\tt libqdf.so}.
All function declarations are available in {\tt qdf.h}
There are three kinds of helpers
\be
\item readers or getters --- Section~\ref{qdf_read_helpers}
\item writers or setters --- Section~\ref{qdf_write_helpers}
\item makers --- Section~\ref{qdf_make_helpers}
  \ee

\subsection{Read only helper functions}
\label{qdf_read_helpers}
Table~\ref{tbl_qdf_read_helpers} lists helper functions that have been provided to
you. They take as an argument an immutable pointer to \qdf\ and do {\bf not}
modify the location pointed to.
\begin{table}
  \centering 
\begin{tabular}{|l|l|l|l|} \hline \hline 
  {\bf Return} & {\bf Name} & {\bf Args} \\ \hline  \hline
  bool & \chk & ( const \qdf\ * const x) \\ \hline 
  \qt & \getq & ( const \qdf\ * const x) \\ \hline 
  \jt & \getj & ( const \qdf\ * const x) \\ \hline 
  uint32\_t  & \getn & ( const \qdf\ * const x) \\ \hline 
  const char * & \getrv & ( const \qdf\ * const x) \\ \hline 
  const bool * & \getrnn & ( const \qdf\ * const x) \\ \hline 
  uint32\_t  & \getnk & ( const \qdf\ * const x) \\ \hline 
\hline
\end{tabular}
\caption{List of read only helper functions}
\label{tbl_qdf_read_helpers}
\end{table}

\subsubsection{\chk}
Returns {\tt true} if x is syntactically valid; {\tt false}, otherwise

\subsubsection{\getq}
Returns \qt  of x 

\subsubsection{\getj}
Returns \jt of x 

\subsubsection{\getn}
If \jt of x is {\tt j\_array}, then returns length of array; else, returns 0.

\subsubsection{\getrv}
\bi
\item If \jt of x is {\tt j\_array}, then returns {\tt NULL}.
\item If \qt of x is Q0, then returns {\tt NULL}.
\item Else, returns a pointer to the \(0^{th}\) element of the array. 
  \ei
The data pointed to cannot be modified. 

\subsubsection{\getrnn}
\bi
\item If \jt of x is {\tt j\_array}, then returns {\tt NULL}.
\item If \qt of x is Qerr or Q0, then returns {\tt NULL}.
\item If x does not have null values, then return {\tt NULL}.
  \ei
Else, returns a pointer to the \(0^{th}\) element of the bool array, nnx,
such that 
\begin{displaymath}
  nnx[i] \Rightarrow x[i] \neq \bot
\end{displaymath}
\begin{displaymath}
  \neg nnx[i] \Rightarrow x[i] = \bot
\end{displaymath}
The data pointed to cannot be modified. 

\subsubsection{\getnk}
If \jt x is {\tt j\_object}, then returns number of keys; else, returns 0.


\subsection{Maker helper functions}
\label{qdf_make_helpers}
Table~\ref{tbl_qdf_make_helpers} lists helper functions that 
have been provided to you to create or make a \qdf.
\be
\item The first argument is always a pointer to \qdf, the elements of which will
  be set by this function
\item The return value is an integer status which is 0 for success 
  and non-zero for failure. 
\item They may have other arguments as well. 
  \ee

\begin{table}
  \centering 
\begin{tabular}{|l|l|l|l|} \hline \hline 
  {\bf Name} & {\bf Arguments} \\ \hline  \hline
  \mknnUA & \qdf x\\ \hline 
         & \qt  q \\ \hline 
         & \uint  n \\ \hline 
  \mkUA &  \qdf x\\ \hline 
         & \qt  q \\ \hline 
         & \uint  n \\ \hline 
\hline
\end{tabular}
\caption{List of maker helper functions}
\label{tbl_qdf_make_helpers}
\end{table}

\subsection{\mkUA}
\label{mkUA}

Creates an empty QDF meant to store a uniform array that does 
{\bf not} have null values. 


\subsection{\mknnUA}
\label{mkUA}

Creates an empty QDF meant to store a uniform array that 
{\bf does} have null values. 

\subsection{Writable helper functions}
\label{qdf_write_helpers}
Table~\ref{tbl_qdf_write_helpers} 
lists helper functions that have been provided to you to modify the elements of a variable of type \qdf. 
\be
\item The first argument is always a pointer to \qdf\ which can be used to
modify the location pointed to. 
\item The return value is an integer status which is 0 for success 
  and non-zero for failure. 
\item They may have other arguments as well. 
  \ee
\begin{table}
  \centering 
\begin{tabular}{|l|l|l|l|} \hline \hline 
  {\bf Name} & Argument 2 & Argument 3 \\ \hline 
  TODO & TODO & TODO \\ \hline
  TODO & TODO & TODO \\ \hline
\hline
\end{tabular}
\caption{List of read only helper functions}
\label{tbl_qdf_write_helpers}
\end{table}

